A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-13 jour/jour.tex
more minor edits

When instructed, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we may construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall our construction of $\ValC$ requires picking values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To pick values $\MU {\Varec{x}}.\; t$, we use a fair enumerator for each (co)datatype $\tau$ which enumerates the values
of type $\tau$ in a normal interpretation, which are closed normal $\MU$-terms.
Concretely, we enumerate a stream of \emph{every} $\MU$-term of our signature $\Sigma$ in a fair manner,
discarding whose that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$,
which we do not explicitly construct.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
Instead, to check if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is alpha-equivalent to a closed-term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently checked by matching the term $\MU {\Varec{x}}.\; t$ with the term $\MU {\Varec{y}}.\; u$.

@@

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas may entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
When assuming that $\ty{u}$ has infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the $\ty{stream}_{\,\ty{u}}$ codatatype %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$.
In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.

>>>>

When instructed to do so, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we %may
construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall that our construction of $\ValC$ requires picking values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To pick values $\MU {\Varec{x}}.\; t$, we use a fair enumerator for each (co)datatype $\tau$,
which lists all the values
of type $\tau$ in a normal interpretation. %, which are closed normal $\MU$-terms.

%Concretely,
For codatatypes, we enumerate a stream of \relax{every} $\MU$-term of our signature $\Sigma$ in a fair manner,
discarding those that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
This set does not need to be explicitly constructed.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To determine if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is $\alpha$-equivalent to a closed term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently achieved by matching the term $\MU {\Varec{x}}.\; t$ with %the term
$\MU {\Varec{y}}.\; u$.

@@

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas may entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
When assuming that $\ty{u}$ has infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the $\ty{stream}_{\,\ty{u}}$ codatatype %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.
